The Talking Heads Experiment

### 2015-04-27 chap3/chap3.tex
fixed remaining typos

Similar behaviour systems and networks have
been experimented with 
for more complex tasks and it is actually the way that 
the Mars rover discussed in Chapter 1 works.\footnote{At 
the risk of simplifying, we can say that the early 
work in cybernetics (such as that of \cite{Braitenberg:1984}) 
has focused on this subsymbolic layer
and that {\it Artificial Intelligence}, as a research field
emerging in the late 1950's, has focused on the 
symbolic layer. Some researchers involved in a bottom up 
approach to Artificial Intelligence strongly refocused
on the subsymbolic layer. One of the most
vocal advocates of this position is \cite{Brooks:1992}.
Obviously we need the two, 
but bridging the gap is a non-trivial problem, sometimes known 
as the grounding problem. See: \cite{Harnad:1994}. }
However, the gap between the continuous dynamics of sensori-motor
intelligence and cognition remains unbridged. 
It is possible for {\itshape us} to see 
structure in the data but this structure is not perceived
nor used for control by the robot itself. 
The robot does not segment nor categorise
reality. It does not ``know'' that it is moving left or right 
and therefore cannot communicate this information to another robot. 
All processing remains at the analog continuous
level. However because it remains at this subsymbolic level, it is doubtful
whether we can ever hope to bootstrap cognitive
intelligence simply by adding more of these dynamical 
systems.\footnote{
\cite{Marr:1982} remains a classical reference outlining
the features that can be extracted, the
principled algorithms for doing it, and possible 
neurophysiological models. These are the main research
topics that still dominate research in vision. See 
\cite{Ullman:1996} for an overview.}
The difference between behavioural intelligence 
and cognitive intelligence resides in an additional 
layer of processing which is no longer continuous and analog 
but discrete and symbolic. How this second symbolic 
layer could be formed but at the same time remain 
grounded in the analog sensori-motor layer is one 
of the main research topics addressed in this book.

@@

The visual layer of the Talking Heads only extracts 
patches and edges which each leads to one way 
of segmenting the scene. 
Segmenting based on patches attempts to aggregate 
those parts of the image into patches (a process
called {\it region growing}) that have more or less the same 
colour. More or less the same is of course a relative
notion and larger or smaller segments will be found
depending on the thresholds that are used for 
deciding whether a colour is similar or not. Segments
that are too small are not considered further. 
Region growing starts by comparing for each pixel how similar it is to 
neighbouring pixels. Similar pixels
are grouped as a patch. In the next step, each patch
is examined to see whether it can be merged with
a neighbouring patch, and so on recursively until 
patches cannot be combined further. Small patches or individual pixels that 
stand on their own are included as part of a larger
patch so that we get sufficiently broad patches.

@@

\is{sieve architecture}Segmentation exemplifies a dual kind of processing that 
we will encounter again and again as we focus on the 
different layers of the cognitive 
architecture (\figref{f:sieve}). 
Various possible solutions are generated, 
expanded and combined in parallel. The possible solutions enter
into competition until globally coherent
solutions emerge, which are ranked and handed to the 
next layer of processing. Often 
a solution does not emerge or multiple solutions are
equally valid in which case constraints from 
expectations or from the further processing of partial
solutions must flow down to influence earlier processing, 
which can be implemented as a {\it re-entry} of some solutions
back into the previous layer. For example, hearing a word may 
stimulate the expectation that a certain category is
relevant, which in turn may stimulate the computation
of certain features and influence how the image is 
segmented. A speaker may have to choose between alternative
segmentations and categorisations depending on whether 
the conceptualisation can be succinctly and accurately 
lexicalised and possible lexicalisations will only be 
acceptable when they can be integrated in a complete sentence.

@@

The two way flow of constraints (from perception to 
high level cognitive processing and from high level 
cognitive processing to perception) suggests 
that the brain is best thought of as a massively parallel, 
densely connected processing system, in which multiple 
partial solution structures float up or down, gathering strength 
or weakening with new information entering the system. 
This contrasts with the view that \enlargethispage{1\baselineskip}
information is processed in a serial step-by-step fashion
through tightly compartmentalised modules.\footnote{
See \cite{Fodor:1983}. This book discusses a modular, sequential 
information processing architecture for cognition. A 
non-modular, parallel view 
is sketched in: \cite{Minsky:1985}. 
A more realistic neurophysiological model similar to the 
one underlying the {\it sieve architecture} we have used is
discussed by \cite{Edelman:1987}.} It is true
that we have the illusion that there 
is some {\itshape homunculus}, a little man, which sees a single coherent 
picture of reality and finds quickly the right way to 
verbalise or interpret this picture, but closer examination 
of the actual informational requirements of each subprocess
shows that this can never work. Constraints must flow in all 
directions, because the sensory data has not enough information to allow 
a unique segmentation, and categorisation and language is 
too full of ambiguities to allow a straightforward linear
interpretation.

@@

\is{conceptual spaces}The sensory channels used in the experiments have
deliberately been kept to a minimum so that we can follow 
in detail the ontological and lexical dynamics, which is a 
non-trivial matter as I will demonstrate in Chapters 6 and 7.
Obviously to get a richer lexicon, more channels need to be made 
available to the agents. These channels can often be 
grouped to form a {\it conceptual space}.\footnote{See \cite{Gardenfors:1999}. 
More cognitive oriented spaces, further removed 
from perception, are discussed in \cite{Fauconnier:1994}.} One of the 
best known examples is the colour space formed by the
yellow-blue, red-green, black-white, saturation, and brightness channels.

@@

\item More properties of the shape of segments can be computed.
I have already introduced the \textsc{ratio} channel, which is the area of the 
segment divided by the area of the bounding box, giving an 
indication how rectangular a segment is. This can be generalised
by using an ellips around a shape so that we can calculate
convexity. We can then also calculate elongation (by calculating
the principal axes of the ellipse). 
\end{enumerate}

@@

\is{perceptual constancy}Real world sensory data is remarkably volatile due to the 
high variation and constant change of 
our physical environment, nevertheless human beings have 
the illusion of constancy. 
For example, the colour of an object is determined 
by the wavelength of the light reflected by its surface. 
For a long time, it 
was thought that colour was an intrinsic property of a 
surface, and that we therefore experience the colour of 
an object as constant, but psycho-experimental evidence has shown 
that this is not true. When we look at a surface in 
isolation which reflects light between 
430 and 470 nanometers, we experience 
the colour blue. As a result, one might assume
that blue can be equated
with the experience of light in this particular wavelength 
region. However, if the same surface is perceived
in a broader context and the light conditions 
are changed (so that objectively they now 
reflect light in a different spectral region), they still 
appear as blue! What should objectively 
appear green according to the measured wavelength 
values is still experienced
as blue. This explains why we see colours
as relatively constant even if the light conditions
are changing (for example from broad daylight to 
evening light) which is obviously extremely useful 
for dealing with rich environments. But it 
implies that colour is {\itshape not}
an intrinsic property that can be measured objectively 
with a light meter. It is actively mapped onto reality 
as a result of complex processing which takes the 
context into account.\footnote{
See the discussion about colour 
in: \cite{Varela:1991}}

>>>>

Similar behaviour systems and networks have
been experimented with 
for more complex tasks and it is actually the way that 
the Mars rover discussed in Chapter 1 works.\footnote{At 
the risk of simplifying, we can say that the early 
work in cybernetics (such as that of \cite{Braitenberg:1984}) 
has focused on this subsymbolic layer
and that {\itshape Artificial Intelligence}, as a research field
emerging in the late 1950's, has focused on the 
symbolic layer. Some researchers involved in a bottom up 
approach to Artificial Intelligence strongly refocused
on the subsymbolic layer. One of the most
vocal advocates of this position is \cite{Brooks:1992}.
Obviously we need the two, 
but bridging the gap is a non-trivial problem, sometimes known 
as the grounding problem. See: \cite{Harnad:1994}. }
However, the gap between the continuous dynamics of sensori-motor
intelligence and cognition remains unbridged. 
It is possible for {\itshape us} to see 
structure in the data but this structure is not perceived
nor used for control by the robot itself. 
The robot does not segment nor categorise
reality. It does not ``know'' that it is moving left or right 
and therefore cannot communicate this information to another robot. 
All processing remains at the analog continuous
level. However because it remains at this subsymbolic level, it is doubtful
whether we can ever hope to bootstrap cognitive
intelligence simply by adding more of these dynamical 
systems.\footnote{
\cite{Marr:1982} remains a classical reference outlining
the features that can be extracted, the
principled algorithms for doing it, and possible 
neurophysiological models. These are the main research
topics that still dominate research in vision. See 
\cite{Ullman:1996} for an overview.}
The difference between behavioural intelligence 
and cognitive intelligence resides in an additional 
layer of processing which is no longer continuous and analog 
but discrete and symbolic. How this second symbolic 
layer could be formed but at the same time remain 
grounded in the analog sensori-motor layer is one 
of the main research topics addressed in this book.

@@

The visual layer of the Talking Heads only extracts 
patches and edges which each leads to one way 
of segmenting the scene. 
Segmenting based on patches attempts to aggregate 
those parts of the image into patches (a process
called {\itshape region growing}) that have more or less the same 
colour. More or less the same is of course a relative
notion and larger or smaller segments will be found
depending on the thresholds that are used for 
deciding whether a colour is similar or not. Segments
that are too small are not considered further. 
Region growing starts by comparing for each pixel how similar it is to 
neighbouring pixels. Similar pixels
are grouped as a patch. In the next step, each patch
is examined to see whether it can be merged with
a neighbouring patch, and so on recursively until 
patches cannot be combined further. Small patches or individual pixels that 
stand on their own are included as part of a larger
patch so that we get sufficiently broad patches.

@@

\is{sieve architecture}Segmentation exemplifies a dual kind of processing that 
we will encounter again and again as we focus on the 
different layers of the cognitive 
architecture (\figref{f:sieve}). 
Various possible solutions are generated, 
expanded and combined in parallel. The possible solutions enter
into competition until globally coherent
solutions emerge, which are ranked and handed to the 
next layer of processing. Often 
a solution does not emerge or multiple solutions are
equally valid in which case constraints from 
expectations or from the further processing of partial
solutions must flow down to influence earlier processing, 
which can be implemented as a {\itshape re-entry} of some solutions
back into the previous layer. For example, hearing a word may 
stimulate the expectation that a certain category is
relevant, which in turn may stimulate the computation
of certain features and influence how the image is 
segmented. A speaker may have to choose between alternative
segmentations and categorisations depending on whether 
the conceptualisation can be succinctly and accurately 
lexicalised and possible lexicalisations will only be 
acceptable when they can be integrated in a complete sentence.

@@

The two way flow of constraints (from perception to 
high level cognitive processing and from high level 
cognitive processing to perception) suggests 
that the brain is best thought of as a massively parallel, 
densely connected processing system, in which multiple 
partial solution structures float up or down, gathering strength 
or weakening with new information entering the system. 
This contrasts with the view that \enlargethispage{1\baselineskip}
information is processed in a serial step-by-step fashion
through tightly compartmentalised modules.\footnote{
See \cite{Fodor:1983}. This book discusses a modular, sequential 
information processing architecture for cognition. A 
non-modular, parallel view 
is sketched in: \cite{Minsky:1985}. 
A more realistic neurophysiological model similar to the 
one underlying the {\itshape sieve architecture} we have used is
discussed by \cite{Edelman:1987}.} It is true
that we have the illusion that there 
is some {\itshape homunculus}, a little man, which sees a single coherent 
picture of reality and finds quickly the right way to 
verbalise or interpret this picture, but closer examination 
of the actual informational requirements of each subprocess
shows that this can never work. Constraints must flow in all 
directions, because the sensory data has not enough information to allow 
a unique segmentation, and categorisation and language is 
too full of ambiguities to allow a straightforward linear
interpretation.

@@

\is{conceptual spaces}The sensory channels used in the experiments have
deliberately been kept to a minimum so that we can follow 
in detail the ontological and lexical dynamics, which is a 
non-trivial matter as I will demonstrate in Chapters 6 and 7.
Obviously to get a richer lexicon, more channels need to be made 
available to the agents. These channels can often be 
grouped to form a {\itshape conceptual space}.\footnote{See \cite{Gardenfors:1999}. 
More cognitive oriented spaces, further removed 
from perception, are discussed in \cite{Fauconnier:1994}.} One of the 
best known examples is the colour space formed by the
yellow-blue, red-green, black-white, saturation, and brightness channels.

@@

\item More properties of the shape of segments can be computed.
I have already introduced the \textsc{ratio} channel, which is the area of the 
segment divided by the area of the bounding box, giving an 
indication how rectangular a segment is. This can be generalised
by using an ellipse around a shape so that we can calculate
convexity. We can then also calculate elongation (by calculating
the principal axes of the ellipse). 
\end{enumerate}

@@

\is{perceptual constancy}Real world sensory data is remarkably volatile due to the 
high variation and constant change of 
our physical environment, nevertheless human beings have 
the illusion of constancy. 
For example, the colour of an object is determined 
by the wavelength of the light reflected by its surface. 
For a long time, it 
was thought that colour was an intrinsic property of a 
surface, and that we therefore experience the colour of 
an object as constant, but psycho-experimental evidence has shown 
that this is not true. When we look at a surface in 
isolation which reflects light between 
430 and 470 nanometres, we experience 
the colour blue. As a result, one might assume
that blue can be equated
with the experience of light in this particular wavelength 
region. However, if the same surface is perceived
in a broader context and the light conditions 
are changed (so that objectively they now 
reflect light in a different spectral region), they still 
appear as blue! What should objectively 
appear green according to the measured wavelength 
values is still experienced
as blue. This explains why we see colours
as relatively constant even if the light conditions
are changing (for example from broad daylight to 
evening light) which is obviously extremely useful 
for dealing with rich environments. But it 
implies that colour is {\itshape not}
an intrinsic property that can be measured objectively 
with a light meter. It is actively mapped onto reality 
as a result of complex processing which takes the 
context into account.\footnote{
See the discussion about colour 
in: \cite{Varela:1991}}

### 2015-04-27 chap4/chap4.tex
hyphenations, footnotes, ...
fixed remaining typos

In recent decades, various designs for neural network models
have been proposed that reflect this empiricist stance.\footnote{
The first neural network models emerged in the 
fifties from the work of neurologists and computer 
scientists like Donald Hebb, McCullogh and Pitts, 
Rosenblatt, and others. There was a strong first wave of 
enthusiasm in the sixties, as illustrated for 
example in: \cite{Minsky:1968}. 
A second wave developed in the mid-eighties when 
new more powerful network architectures were discovered
that could handle intermediary representations and 
later on temporal structures (see the overview 
in \cite{Churchland:1992}. }
The input nodes of these networks receive data
from sensory channels of the sort discussed in the 
previous chapter. They are connected to higher
level nodes which use the data to decide whether
a category applies. The connections are weighted and 
a positive output signal is produced when 
the weighted sum of the inputs 
exceeds a certain threshold. Thus the networks
exhibit some of the flexibility, tolerance
to variation, and context-dependence seen in human
categorisation. The weights and thresholds are 
learned by propagating back errors in categorisation. 
If a node makes a positive identification when 
it should not have done so, the weights of the 
incoming connections are lowered and the threshold is
increased, so that it is less likely that the threshold
will be exceeded again for the same situation the next time around. 
Conversely, if the network makes a negative identification where
it should have made a positive one, the weights 
of the incoming connections are increased 
and the threshold lowered. It is known 
through mathematical proofs that such networks indeed
stabilise on reliable categorisations, if the 
environment remains sufficiently constant.\footnote{
This type of neural networks and some of its
main variants are discussed at length in the classical 
textbook by \cite{Rumelhart:1986}.}
These inductive neural networks therefore
constitute ther first serious proposal for bridging the gap between
sensory signals and categories and for explaining the 
origins and acquisition of categories.

@@

Empiricists have had considerable
success in coming up with inductive learning mechanisms. 
They have even been demonstrated on autonomous robots 
in direct interaction with the environment.
At the same time, the learning mechanisms proposed so far turn out
to be very fragile.\footnote{
The weakness of traditional connectionist networks
with respect to speed and resilience
are discussed at length in \cite{Quartz:1997}.}
The human experimenter has to carefully 
set up the architecture of the network (the number of
nodes, the number of layers of nodes, and how they are
connected), tune the learning
parameters, and supply just the right set of test cases. 
Performance may degrade when too many cases are seen. 
Even worse, when new cases are supplied that require
a revision of categories already learned, a substantial
portion of the earlier cases must be resubmitted to 
retrain the network. All this contradicts
the robustness and open-ended extensibility
of human categorisation. In addition, 
the learning mechanisms proposed have been
slow to consistently aquire categories. A large number of cases
are required and often 
cases must go through multiple iterations. Moreover 
any inductive method is a slave of
the data. A category will only become reliably recognised
when it is statistically significant.

@@

\is{selectionism}The difficulties encountered with the empiricist
{\itshape and} rationalist points of view make it worthwhile
to explore alternative solutions. The one I propose
and further explore in this book 
has been inspired by two key principles 
from biology. The first principle is
that of selectionism. It requires 
a growth process in the agents that generates possible 
structures, even in the absence of examples, and a pruning
process that removes those
that are irrelevant. The growth and pruning
process is assumed to be biologically given
but not the categories that result from it.  
The second principle that \is{interactionism}
I will adopt is interactionism, put forward by biologists 
to understand how genetic influences {\itshape and} environmental 
impact cooperate to shape an organism. Interactionism 
was first suggested by Piaget (originally a biologist)
to explain the growth of mental capacity in the child. 
His numerous experiments show a gradual progressive construction of 
increasingly more complex ways of categorising and 
conceptualising reality. The child encounters situations that 
can be assimilated, and thus cause entrenchment of
existing structures, as well 
as situations that cannot be handled and require 
the child to accomodate with new 
constructions or reorganisations.\footnote{
The work of Waddington is typical for the 
interactionist approach to development, see \cite{Waddington:1975}. 
Such a view does not necessarily imply that there is 
a pre-determined course of development.
Piaget emphasised a progressive, dynamical view on 
development \cite{Piaget:1985}. 
His work arose prior to the detailed
computational modeling that is now common place in 
cognitive science and there is still an enormous 
work left to demonstrate operationalisations of these
insights. See for a more recent discussion of 
the issues: \cite{Thelen:1994}.}

@@

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [\textsc{gray}-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [\textsc{vpos} 0.0–05]
(lower), and [\textsc{hpos} 0.0–0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared 
to shape 0 and 2.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & Saliency\\ \midrule
\textsc{hpos} & 0.27 & 0.16 & 0.11\\ 
\textsc{vpos} & 0.20 & 0.20 & 0.0\\ 
\textsc{height} & 0.15 & 0.15 & 0.0\\ 
\textsc{width} & 0.10 & 0.11 & 0.01\\ 
\textsc{area} & 0.10 & 0.10 & 0.0\\ 
R & 0.23 & 0.25 & 0.02\\ 
G & 0.32 & 0.34  & 0.02\\ 
B & 0.63 & 0.65 & 0.02\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-real}Sensory data from a real world scene with segmentation shown in \figref{f:plate-10}.}
\end{center}
\end{table}
Clearly \textsc{hpos} is the most salient channel and should be 
prefered by the agent. When performing context scaling, 
the two values for \textsc{hpos} are drawn apart with 1.0 for object-0 and 
0.0 for object-1 so that the category [\textsc{hpos} 0.5–1.0] (to the 
right) easily distinguishes the topic (object-0) from
object-1. The game is reported as follows: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [HPOS 0.5–1.0]
\end{verbatim}
This example illustrates well why the categorisation 
of the Talking Heads is so robust and why agents 
often share the same conceptualisation even if the details  
of their raw perception is quite different. 
The saliency factor helps to focus the agents on those
aspects of the scene that stand out. There is 
an enormous reduction of variation, first by scaling then by 
the categorisation process itself.

@@

Growth needs to be balanced by pruning. Pruning simply means
that a categoriser and thus its pending branches is 
cut away. There are again two
issues: (1) which nodes should be pruned and
(2) when pruning should take place. 
Obviously the score of a category should play a role
in deciding whether it should be pruned. 
categorisers that have not been used very much or have a low
success rate are prime candidates for removal, unless
any of their subcategorisers has a high score. The 
monitoring of use and success already played a role in 
determining which category should be prefered, so this
information is available to decide on pruning as well.

@@

\is{discriminatory success}The Discrimination Game is a dynamical system.\footnote{
The theory of complex dynamical systems, which is well 
developed in the natural sciences, provides the 
theoretical foundation for studying the Discrimination
Game. For a general introduction, see \cite{Peitgen:1992}.} A repertoire of
categories emerges in an agent gradually as an attractor of the
growth and pruning dynamics coupled to 
the environment. If growth is strictly proportional to failure
and there is no pruning, a point attractor
is reached as soon as the repertoire
is adequate, i.e. as soon 
as the agent consistently has success
for all the possible cases it encountered. However, 
as soon as the environment or the sensory capabilities of 
the agent change, in other words when new types of 
figures appear or when new sensory channels become available
to the agent, we expect that the repertoire of categories starts
expanding again. This could be seen as an illustration  of the
assimilation-accomodation dynamics envisioned by Piaget.

@@

The main conclusion of this chapter is that a selectionist
approach to the origins of categories
is theoretically and practically feasable. I have 
defined a growth and pruning dynamics which leads to 
an adequate repertoire for 
discriminating one object from the others in the 
same context and I showed that the repertoire is
continuously adapted when the environment changes.

>>>>

In recent decades, various designs for neural network models
have been proposed that reflect this empiricist stance.\footnote{
The first neural network models emerged in the 
fifties from the work of neurologists and computer 
scientists like Donald Hebb, McCullogh and Pitts, 
Rosenblatt, and others. There was a strong first wave of 
enthusiasm in the sixties, as illustrated for 
example in: \cite{Minsky:1968}. 
A second wave developed in the mid-eighties when 
new more powerful network architectures were discovered
that could handle intermediary representations and 
later on temporal structures (see the overview 
in \cite{Churchland:1992}. }
The input nodes of these networks receive data
from sensory channels of the sort discussed in the 
previous chapter. They are connected to higher
level nodes which use the data to decide whether
a category applies. The connections are weighted and 
a positive output signal is produced when 
the weighted sum of the inputs 
exceeds a certain threshold. Thus the networks
exhibit some of the flexibility, tolerance
to variation, and context-dependence seen in human
categorisation. The weights and thresholds are 
learned by propagating back errors in categorisation. 
If a node makes a positive identification when 
it should not have done so, the weights of the 
incoming connections are lowered and the threshold is
increased, so that it is less likely that the threshold
will be exceeded again for the same situation the next time around. 
Conversely, if the network makes a negative identification where
it should have made a positive one, the weights 
of the incoming connections are increased 
and the threshold lowered. It is known 
through mathematical proofs that such networks indeed
stabilise on reliable categorisations, if the 
environment remains sufficiently constant.\footnote{
This type of neural networks and some of its
main variants are discussed at length in the classical 
textbook by \cite{Rumelhart:1986}.}
These inductive neural networks therefore
constitute their first serious proposal for bridging the gap between
sensory signals and categories and for explaining the 
origins and acquisition of categories.

@@

Empiricists have had considerable
success in coming up with inductive learning mechanisms. 
They have even been demonstrated on autonomous robots 
in direct interaction with the environment.
At the same time, the learning mechanisms proposed so far turn out
to be very fragile.\footnote{
The weakness of traditional connectionist networks
with respect to speed and resilience
are discussed at length in \cite{Quartz:1997}.}
The human experimenter has to carefully 
set up the architecture of the network (the number of
nodes, the number of layers of nodes, and how they are
connected), tune the learning
parameters, and supply just the right set of test cases. 
Performance may degrade when too many cases are seen. 
Even worse, when new cases are supplied that require
a revision of categories already learned, a substantial
portion of the earlier cases must be resubmitted to 
retrain the network. All this contradicts
the robustness and open-ended extensibility
of human categorisation. In addition, 
the learning mechanisms proposed have been
slow to consistently acquire categories. A large number of cases
are required and often 
cases must go through multiple iterations. Moreover 
any inductive method is a slave of
the data. A category will only become reliably recognised
when it is statistically significant.

@@

\is{selectionism}The difficulties encountered with the empiricist
{\itshape and} rationalist points of view make it worthwhile
to explore alternative solutions. The one I propose
and further explore in this book 
has been inspired by two key principles 
from biology. The first principle is
that of selectionism. It requires 
a growth process in the agents that generates possible 
structures, even in the absence of examples, and a pruning
process that removes those
that are irrelevant. The growth and pruning
process is assumed to be biologically given
but not the categories that result from it.  
The second principle that \is{interactionism}
I will adopt is interactionism, put forward by biologists 
to understand how genetic influences {\itshape and} environmental 
impact cooperate to shape an organism. Interactionism 
was first suggested by Piaget (originally a biologist)
to explain the growth of mental capacity in the child. 
His numerous experiments show a gradual progressive construction of 
increasingly more complex ways of categorising and 
conceptualising reality. The child encounters situations that 
can be assimilated, and thus cause entrenchment of
existing structures, as well 
as situations that cannot be handled and require 
the child to accommodate with new 
constructions or reorganisations.\footnote{
The work of Waddington is typical for the 
interactionist approach to development, see \cite{Waddington:1975}. 
Such a view does not necessarily imply that there is 
a pre-determined course of development.
Piaget emphasised a progressive, dynamical view on 
development \cite{Piaget:1985}. 
His work arose prior to the detailed
computational modeling that is now common place in 
cognitive science and there is still an enormous 
work left to demonstrate operationalisations of these
insights. See for a more recent discussion of 
the issues: \cite{Thelen:1994}.}

@@

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [\textsc{gray}-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [\textsc{vpos} 0.0–05]
(lower), and [\textsc{hpos} 0.0–0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared \enlargethispage{1\baselineskip}
to shape 0 and 2.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & Saliency\\ \midrule
\textsc{hpos} & 0.27 & 0.16 & 0.11\\ 
\textsc{vpos} & 0.20 & 0.20 & 0.0\\ 
\textsc{height} & 0.15 & 0.15 & 0.0\\ 
\textsc{width} & 0.10 & 0.11 & 0.01\\ 
\textsc{area} & 0.10 & 0.10 & 0.0\\ 
R & 0.23 & 0.25 & 0.02\\ 
G & 0.32 & 0.34  & 0.02\\ 
B & 0.63 & 0.65 & 0.02\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-real}Sensory data from a real world scene with segmentation shown in \figref{f:plate-10}.}
\end{center}
\end{table}
Clearly \textsc{hpos} is the most salient channel and should be 
preferred by the agent. When performing context scaling, 
the two values for \textsc{hpos} are drawn apart with 1.0 for object-0 and 
0.0 for object-1 so that the category [\textsc{hpos} 0.5–1.0] (to the 
right) easily distinguishes the topic (object-0) from
object-1. The game is reported as follows: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [HPOS 0.5–1.0]
\end{verbatim}
This example illustrates well why the categorisation 
of the Talking Heads is so robust and why agents 
often share the same conceptualisation even if the details  
of their raw perception is quite different. 
The saliency factor helps to focus the agents on those
aspects of the scene that stand out. There is 
an enormous reduction of variation, first by scaling then by 
the categorisation process itself.

@@

Growth needs to be balanced by pruning. Pruning simply means
that a categoriser and thus its pending branches is 
cut away. There are again two
issues: (1) which nodes should be pruned and
(2) when pruning should take place. 
Obviously the score of a category should play a role
in deciding whether it should be pruned. 
categorisers that have not been used very much or have a low
success rate are prime candidates for removal, unless
any of their subcategorisers has a high score. The 
monitoring of use and success already played a role in 
determining which category should be preferred, so this
information is available to decide on pruning as well.

@@

\is{discriminatory success}The Discrimination Game is a dynamical system.\footnote{
The theory of complex dynamical systems, which is well 
developed in the natural sciences, provides the 
theoretical foundation for studying the Discrimination
Game. For a general introduction, see \cite{Peitgen:1992}.} A repertoire of
categories emerges in an agent gradually as an attractor of the
growth and pruning dynamics coupled to 
the environment. If growth is strictly proportional to failure
and there is no pruning, a point attractor
is reached as soon as the repertoire
is adequate, i.e. as soon 
as the agent consistently has success
for all the possible cases it encountered. However, 
as soon as the environment or the sensory capabilities of 
the agent change, in other words when new types of 
figures appear or when new sensory channels become available
to the agent, we expect that the repertoire of categories starts
expanding again. This could be seen as an illustration  of the
assimilation-accommodation dynamics envisioned by Piaget.

@@

The main conclusion of this chapter is that a selectionist
approach to the origins of categories
is theoretically and practically feasible. I have 
defined a growth and pruning dynamics which leads to 
an adequate repertoire for 
discriminating one object from the others in the 
same context and I showed that the repertoire is
continuously adapted when the environment changes.
